---
title: "braidrm: Package Introduction"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(braidrm)
library(braidReports)
library(tidyverse)
```

## Introduction

This package contains most of the essential code for fitting and evaluating
response surfaces of the Bivariate Response to Additive Interacting Doses,
or BRAID, model.  This article walks through some of the most basic
functionality of the package, including evaluating, inverting, and fitting
BRAID response surfaces.  Further details, exceptions and functions can be
explored in other vignettes. *Note: this package makes use of both the*
`braidReports` *package for visualization and the* `tidyverse` *package to*
*make the code more easily readable. Neither is a required dependency of the*
`braidrm` *package, though* `braidReports` *is strongly suggested.*

## The BRAID Model {-}

The BRAID model is a parametric model of combined action, which draws its
inspiration (and mathematical form) from the Hill single-agent dose response
model, also often, known as the log-logistic function:

$$
E(D) = E_0 + \frac{E_f - E_0}{1 + \left(\frac{D}{{ID}_M}\right)^{-n}}
$$

In this equation, $D$ is the dose of the compound or agent in question;
${ID}_M$ is the dose of median effect (sometimes referred to as the $EC_{50}$);
$n$ is the "Hill slope" or "sigmoidicity" (reflecting how quickly the agent
goes from low effects to high effects as a function of concentration/dose);
$E_0$ is the minimal effect, expected when no agent is present; $E_f$ is the
maximal effect, the asymptotic level expected when the agent is present at
very high concentrations; and $E(D)$ is the predicted effect of the specified
dose according to the model.

In the BRAID model, we have two agents (call them A and B), so we assume that
their individual dose response behaviors are reprsented by the following
equations:

$$
{E_A}(D_A) = E_0 + \frac{E_{f,A} - E_0}{1 + \left(\frac{D}{{ID}_{M,A}}\right)^{-n_a}}
$$

$$
{E_B}(D_B) = E_0 + \frac{E_{f,B} - E_0}{1 + \left(\frac{D}{{ID}_{M,B}}\right)^{-n_b}}
$$

Note that both models use the same parameter $E_0$, as the expected effect when
neither drug is present should be the same.  The BRAID model merges these two
individual behaviors into a single parametric model.  A full treatment of the
equation can be found in `vignette("derivation")`, but a simpler version
(in which the Hill slopes and maximal effects of the drugs are assumed to be
equal) can be written as:

$$
E(D_A,D_B) = E_0 + \frac{E_f - E_0}{1+\left(\frac{D}{{ID}_{M,A}} + \frac{D}{{ID}_{M,B}} + \kappa\sqrt{\frac{D}{{ID}_{M,A}}\frac{D}{{ID}_{M,B}}}\right)^{-n}}
$$

The BRAID model adds a single additional parameter beyond the individual dose
response parameters of the two drugs: the Greek letter $\kappa$, representing
the degree of synergy or antagonism.  When $\kappa=0$, the two doses add
together, behaving as though they were diluted doses of the same drug; this is
classical notion of Loewe additivity.  When $\kappa>0$, if both drugs are 
present, the effective dose is increased above simple addition, the effect is
higher, and the combination exhibits synergy.  When $\kappa<0$, if both drugs
are prsent, the effective dose is *decreased* below simple addition, the effect
is reduced, and the combination ehxibits antagonism.  As $\kappa$ goes higher
(theoretically to infinity), the degree of synergy increases; as $\kappa$ goes
lower (to a minimum of $-2$), the degree of antagonism increases.

## BRAID Surface Examples {-}

We can build an example surface using the function `evalBraidModel()`.  The 
following code creates a parameter vector where ${ID}_{M,A} = {ID}_{M,B} = 1$, 
$n_a = n_b = 3$, $E_0 = 0$, $E_{f,A} = E_{f,B} = E_f = 1$, and $\kappa = 0$ 
(indicating an additive surface), and then generates a response surface for that
parameter vector.

```{r}
additivePar <- c(
	1, # IDMA
	1, # IDMB
	3, # na
	3, # nb
	0, # kappa
	0, # E0
	1, # EfA
	1, # EfB
	1  # Ef
)

concentrations <- seq(0,3,length=51)
surface <- expand_grid(concA=concentrations, concB=concentrations) %>%
	mutate(additive = evalBraidModel(concA,concB,additivePar))

ggplot(surface,aes(x=concA,y=concB)) +
	geom_raster(aes(fill=additive)) +
	scale_fill_distiller(palette="RdYlBu",direction=-1) +
	geom_contour(aes(z=additive),breaks=seq(0.1,0.9,by=0.1),
				 colour="black",linetype=2)+
	labs(x="Concentration A",
		 y="Concentration B",
		 fill="Effect",
		 title="Additive Surface")+
	coord_equal()
```

Another view of the surface can be given by taking slices through the surface at
different levels of one of the drugs.  For example:

```{r}
samples <- c(0,0.5,0.75,0.9,1,1.11,1.33,2)

lines <- expand_grid(concA=concentrations,concB=samples) %>%
	mutate(additive=evalBraidModel(concA,concB,additivePar))

ggplot(lines,aes(x=concA,y=additive))+
	geom_line(aes(color=factor(concB)))+
	labs(x="Concentration A",y="Effect",color="Drug B")

```

Changing the BRAID interaction parameter $\kappa$ leaves the individual dose
response behaviors alone, but alters the way they interact when both are 
present. Note the effect on the contours of the response surface resulting
from positive and negative $\kappa$ values.

```{r}
synergisticPar <- c(
	1, 1, 3, 3,
	2, # kappa > 0 indicates synergy
	0, 1, 1, 1 
)
antagonisticPar <- c(
	1, 1, 3, 3,
	-1, # kappa < 0 indicates antagonism
	0, 1, 1, 1 
)

surface <- surface %>%
	mutate(synergy = evalBraidModel(concA,concB,synergisticPar),
		   antagonism = evalBraidModel(concA,concB,antagonisticPar))

ggplot(surface,aes(x=concA,y=concB)) +
	geom_raster(aes(fill=synergy)) +
	scale_fill_distiller(palette="RdYlBu",direction=-1) +
	geom_contour(aes(z=synergy),breaks=seq(0.1,0.9,by=0.1),
				 colour="black",linetype=2)+
	labs(x="Concentration A",
		 y="Concentration B",
		 fill="Effect",
		 title="Synergistic Surface")+
	coord_equal()

ggplot(surface,aes(x=concA,y=concB)) +
	geom_raster(aes(fill=antagonism)) +
	scale_fill_distiller(palette="RdYlBu",direction=-1) +
	geom_contour(aes(z=antagonism),breaks=seq(0.1,0.9,by=0.1),
				 colour="black",linetype=2)+
	labs(x="Concentration A",
		 y="Concentration B",
		 fill="Effect",
		 title="Antagonistic Surface")+
	coord_equal()

```

## Fitting BRAID Surfaces to Data {-}

``` {r, include=FALSE}
set.seed(20240730)
```

Of course the primary use of the BRAID model, like the Hill model that inspired
it, is to fit (and hence parametrically summarize) experimentally measured data.
For simplicity, we will generate some artificial combination data with a 
moderately synergistic combination.

```{r, warning=FALSE}
measurements <- synergisticExample

ggplot(measurements,aes(x=concA,y=concB)) +
	geom_braid(aes(fill=measure))+
	scale_fill_distiller(palette="RdYlBu",direction=-1) +
	scale_x_log10()+
	scale_y_log10()+
	labs(x="Concentration A",
		 y="Concentration B",
		 fill="Effect",
		 title="Measured Surface")+
	coord_equal()
```

The simplest way to fit a BRAID model to data is with the `braidrm()` function:

```{r}
braidFit <- braidrm(measure ~ concA + concB, measurements,
					model="kappa2", getCIs=FALSE)

summary(braidFit)
```

This "formula" usage of `braidrm()` indicates that we should take our values 
from the data frame `measurements`, and that the response surface to be fit
has the column `measured` containing the modeled effect as a function of
concentrations in `concA` and `concB`.  The value "kappa2" for the parameter
`model` is the default value, and indicates we should fit a standard 8-parameter
BRAID surface in which the two parameters $E_{f,A}$ and $E_{f,B}$ are both fit
and may differ from each other. (Setting the parameter `getCIs` to `FALSE` just
indicates that we do not want to calculate confidence intervals yet.)

Plotting the fitted model we can see that it matches the original data quite
closely:

```{r, warning=FALSE}
measurements <- measurements %>%
	mutate(fit = fitted(braidFit))

ggplot(measurements,aes(x=concA,y=concB)) +
	geom_braid(aes(fill=fit))+
	scale_fill_distiller(palette="RdYlBu",direction=-1) +
	scale_x_log10()+
	scale_y_log10()+
	labs(x="Concentration A",
		 y="Concentration B",
		 fill="Effect",
		 title="Measured Surface")+
	coord_equal()

r2 <- signif(cor(measurements$fit,measurements$measure)^2,3)
ggplot(measurements,aes(x=fit,y=measure)) +
	geom_abline(color="red")+
	geom_linerange(aes(ymin=fit,ymax=measure),linetype=2)+
	geom_point()+
	labs(x="Predicted",
		 y="Measured",
		 title=sprintf("R\u00B2 = %s",r2))+
	coord_equal()
```

Another useful function for fitting BRAID surfaces is `findBestBraid()`, which
tries several different candidate BRAID models and selects the optimal model 
using the Bayesian or Akaike information critieria.  These candidate models
differ in how they treat the minimal and maximal effects, fixing them at
constant default values or equal to each other in some models and letting them 
vary freely in others.  Here we see an example usage of the function in which
the minimal effect default is very close to the best fit, but the default 
maximal effect is not.

```{r}
bestFit <- findBestBraid(measure ~ concA + concB, measurements,
						 defaults = c(0,2), getCIs = TRUE)

summary(bestFit)
```

The summary of the object `bestFit` contains a matrix of bootstrapped
confidence intervals on all the parameters allowed to vary freely in the best
fitting model.  As $E_0$ CI values are `NA` in this array, we know that the best
(read: most parsimonious) model allowed $E_0$ to be fixed at the default value
of 0.  The parameter $E_f$ *is* included, but the parameters $E_{f,A}$ and 
$E_{f,B}$ are not, indicating that the model fixed all maximal effect parameters
equal to a single value and allowed that value to vary (being fit, in this case,
very close to 1).  See the documentation of `findBestBraid()` for more details
on the candidate models and their meaning.

## Inverting a BRAID Model {-}

One last operation that is commonly needed with a response surface is to
determine how much of one drug is required to produce a desired effect in the
presence of another; this amount is sometimes referred to as an inhibitory
concentration (IC) or inhibitory dose (ID).  For consistency, we will use IC to
describe such values.  So the concentration required to have a 75% complete 
effect might be referred to as the ${IC}_{75}$. This can be calculated from a
BRAID surface using the function `invertBraidModel()`.

```{r}

bestPar <- coef(bestFit)

baselineIcValues <- invertBraidModel(DB=0,
									 effect=seq(0.1,0.9,by=0.1),
									 bpar=bestPar)
alteredIcValues <- invertBraidModel(DB=1,
									effect=seq(0.1,0.9,by=0.1),
									bpar=bestPar)

names(baselineIcValues) <- paste0("IC",seq(10,90,by=10))
names(alteredIcValues) <- paste0("IC",seq(10,90,by=10))

baselineIcValues
alteredIcValues
```

From these we see that for the best fit BRAID surface, most of the
concentrations required to produce given effects lie close to 1, typical of a 
dose response behavior with a relatively high Hill slope.  However, in the
presence of a dose of 1 unit of agent B, the IC values are drastically reduced
falling to 0 for 10, 20, 30, and 40% (as 1 unit of agent B alone produces
over 40% effect), but also dropping significantly for other effect levels
due to synergy, including a nearly five-fold reduction in ${IC}_{90}$.
