---
title: "Estimating the IAE"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(braidrm)
library(braidReports)
library(tidyverse)
```

## Introduction

While the parameters of the BRAID model are chosen to carry intuitive meaning
about different aspects of a drug combination's effect, analyzing them
individually often leaves out important aspects of the bigger picture.
Determining whether one combination is more or less synergistic than another may
be interesting scientifically, but in a translational setting that comparison is
far less important than a much simpler question: does the combination work? A
less synergistic, but much more potent combination is translationally far more
promising, even if the lower interaction makes it less scientifically exciting.
In dose response analysis, these more translational evaluations are performed
using aggregate measures like the area-under-the-curve or the therapeutic
window. How might we apply these concepts to a response *surface*? 
*Note: this package makes use of both the*
`braidReports` *package for visualization and the* `tidyverse` *package to*
*make the code more easily readable. Neither is a required dependency of the*
`braidrm` *package, though* `braidReports` *is strongly suggested.*


## The Index of Achievable Efficacy

The primary metric of aggregate activity for the BRAID model is the index
of achievable efficacy, or IAE.  Based on the therapeutic index, the IAE is
a quantitative comparison of the space of achievable doses of a given
combination and the space of doses that produce a desired effect. To see it in
action, consider the following example:

Say we have a additive combination of two drugs.  For simplicity, we'll assume
that both drugs have a dose of median effect of 1 micromolar, a relatively
sharp Hill slope of 3, and maximal effect ot 1.0 (100% efficacy). Suppose 
further that both drugs can safely be administered up to 2 micromolar;
unfortunately, this means that neither drug can cross a threshold of 90% 
efficacy in isolation; but when combined, many dose pairs below the 2 micromolar
threshold achieve the desired effect:

```{r}
additivePar <- c(
	1, 1, 3, 3,
	0, # Additive
	0, 1, 1, 1  
)

concentrations <- seq(0,3,length=51)
surface <- expand_grid(concA=concentrations, concB=concentrations) %>%
	mutate(additive = evalBraidModel(concA,concB,additivePar))

achieve <- data.frame(concA=c(-Inf,2,2),concB=c(2,2,-Inf))

ggplot(surface,aes(x=concA,y=concB)) +
	geom_raster(aes(fill=additive)) +
	scale_fill_distiller(palette="RdYlBu",direction=-1) +
	geom_contour(aes(z=additive),breaks=0.9,colour="black")+
	geom_path(data=achieve,linetype=2,colour="black")+
	labs(x="Concentration A",
		 y="Concentration B",
		 fill="Effect",
		 title="Additive Surface")+
	coord_equal()

```

This simple fact - that combining achievable doses of multiple drugs can reach
levels of efficacy not safely achievable with any compound alone - is the 
primary reason for the extraordinary power and potential of combination therapy.

So how can we quantify this efficacy?  When calculating the therapeutic index
one takes the highest safely achievable dose (the magnitude of achievable doses)
and divides it by the lowest dose that reaches some desired level of effect.
Because every dose below this threshold does *not* reach the desired effect,
this threshold dose can also be viewed as the magnitude of achievable doses that
do not reach the desired effect.  We use this somewhat awkward re-framing 
because it translates much more easily to higher dimensional therapeutic spaces.

So to create an analogous metric in a two-drug combination, we take the total
magnitude of achievable dose-space (in this case everything where both drugs 
are below 2 micomolar, so 4 micromolar squared), and divide it by the space of
doses that do *not* reach the desired effect (in this case 0.9 or 90%). This 
ratio is *close* to being the IAE, but because it still carries the
dimensionality of the two-drug space, we take the square root to produce a value
comparable to a traditional therapeutic index.

All these calculations are carried out for us using the function 
`estimateIAE()`:

```{r}
estimateIAE(
	additivePar,  # A Braid parameter vector
	0.9,          # The desired effect level
	c(2,2)        # The maximum achievable concentrations of both drugs
)
```

This value is not extraordinarily high but it can be affected by the properties
of the response surface, including the interaction parameter $\kappa$. 
Not surprisingly increasing $\kappa$ reduces the space of sub-threshold doses,
and hence increases the IAE:

```{r}
synergisticPar <- c(
	1, 1, 3, 3,
	2, # kappa > 0 indicates synergy
	0, 1, 1, 1 
)

surface <- surface %>%
	mutate(synergy = evalBraidModel(concA,concB,synergisticPar))

ggplot(surface,aes(x=concA,y=concB)) +
	geom_raster(aes(fill=synergy)) +
	scale_fill_distiller(palette="RdYlBu",direction=-1) +
	geom_contour(aes(z=synergy),breaks=0.9,colour="black")+
	geom_path(data=achieve,linetype=2,colour="black")+
	labs(x="Concentration A",
		 y="Concentration B",
		 fill="Effect",
		 title="Additive Surface")+
	coord_equal()

estimateIAE(synergisticPar, 0.9, c(2,2))
```

Introducing antagonism on the other hand inflates the space of subthreshold
doses above the range of achievable dose pairs, resulting in a minimal IAE of
1.

```{r}
antagonisticPar <- c(
	1, 1, 3, 3,
	-1, # kappa < 0 indicates antagonism
	0, 1, 1, 1 
)

surface <- surface %>%
	mutate(antagonism  = evalBraidModel(concA,concB,antagonisticPar))

ggplot(surface,aes(x=concA,y=concB)) +
	geom_raster(aes(fill=antagonism )) +
	scale_fill_distiller(palette="RdYlBu",direction=-1) +
	geom_contour(aes(z=antagonism ),breaks=0.9,colour="black")+
	geom_path(data=achieve,linetype=2,colour="black")+
	labs(x="Concentration A",
		 y="Concentration B",
		 fill="Effect",
		 title="Additive Surface")+
	coord_equal()

estimateIAE(antagonisticPar, 0.9, c(2,2))
```
