---
title: "Flipped BRAID Surfaces"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(braidrm)
library(braidReports)
library(tidyverse)
```

## Introduction

Though versatile, the BRAID response surface model makes certain qualitative 
assumptions about the behavior of the agents modeled.  Specifically, it assumes
that both agents produce a measureable effect and that that effect is in the 
same direction and greater for higher doses.  Yet many potential compound
behaviors cannot be molded to this form.  Fortunately, with some fairly simple
manipulations, the BRAID fitting tools can be used to describe a much wider
range of compound behaviors. *Note: this article makes use of both the* `braidReports` 
*package for visualization and the* `tidyverse` *package to make the code more*
*easily readable. Neither is a required dependency of the* `braidrm` *package,*
*though* `braidReports` *is strongly suggested.*

## Motivation

The code below shows us two views of the same BRAID additive response surface.
The first view, in which concentrations are plotted on a linear scale, is more
effective at showing the qualitative behavior of the surface, in particular
the additive behavior of the two compounds.  The second, in which concentrations
are plotted on a logarithmic scale, is more typical of the view afforded by
combination experiments, and highlights the mathematical behavior of the BRAID
model.  Effectively, the model divides the space of doses into four quadrants,
depending on whether the first dose lies above or below ${ID}_{M,A}$ and whether
the second dose lies above or below ${ID}_{M,B}$.  The values in these four
quadrants approach the four effect parameters ($E_0$, $E_{f,A}$, $E_{f,B}$ and
$E_f$), with the sharpness and shape of the transition governed by $n_a$, $n_b$,
and $\kappa$.

```{r}
linearConcs <- seq(0,3,length=51)
logConcs <- exp(seq(log(0.1),log(10),length=51))

linearSurface <- expand_grid(concA=linearConcs,concB=linearConcs) %>%
	mutate(standard=evalBraidModel(concA,concB,c(1,1,3,3,0,0,1,1,1)))
logSurface <- expand_grid(concA=logConcs,concB=logConcs) %>%
	mutate(standard=evalBraidModel(concA,concB,c(1,1,3,3,0,0,1,1,1)))

ggplot(linearSurface,aes(x=concA,y=concB)) +
	geom_raster(aes(fill=standard)) +
	scale_fill_distiller(palette="RdYlBu",direction=-1) +
	geom_contour(aes(z=standard),breaks=seq(0.1,0.9,by=0.1),
				 colour="black",linetype=2)+
	labs(x="Concentration A",
		 y="Concentration B",
		 fill="Effect",
		 title="Standard Surface (linear)")+
	coord_equal()
ggplot(logSurface,aes(x=concA,y=concB)) +
	geom_raster(aes(fill=standard)) +
	scale_fill_distiller(palette="RdYlBu",direction=-1) +
	geom_contour(aes(z=standard),breaks=seq(0.1,0.9,by=0.1),
				 colour="black",linetype=2)+
	scale_x_log10()+
	scale_y_log10()+
	labs(x="Concentration A",
		 y="Concentration B",
		 fill="Effect",
		 title="Standard Surface (logarithmic)")+
	coord_equal()

```

If we were to negate the Hill slope $n_a$ in the BRAID equation, the effect 
would be to treat dose pairs below ${ID}_{M,A}$ like dose pairs above
${ID}_{M,B}$, and vice versa.  In essence, this flips the response surface
along the drug A axis, around the center ${ID}_{M,A}$; this is implemented
in the `braidrm` package using the `evalFlippedBraidModel()` function.

```{r}
linearSurface <- linearSurface %>%
	mutate(protective=evalFlippedBraidModel(concA,concB,
											c(1,1,3,3,0,1,0,1,1),
											flip="A"))
logSurface <- logSurface %>%
	mutate(protective=evalFlippedBraidModel(concA,concB,
											c(1,1,3,3,0,1,0,1,1),
											flip="A"))

ggplot(logSurface,aes(x=concA,y=concB)) +
	geom_raster(aes(fill=protective)) +
	scale_fill_distiller(palette="RdYlBu",direction=1) +
	geom_contour(aes(z=protective),breaks=seq(0.1,0.9,by=0.1),
				 colour="black",linetype=2)+
	scale_x_log10()+
	scale_y_log10()+
	labs(x="Concentration A",
		 y="Concentration B",
		 fill="Effect",
		 title="Protective Surface (logarithmic)")+
	coord_equal()
ggplot(linearSurface,aes(x=concA,y=concB)) +
	geom_raster(aes(fill=protective)) +
	scale_fill_distiller(palette="RdYlBu",direction=1) +
	geom_contour(aes(z=protective),breaks=seq(0.1,0.9,by=0.1),
				 colour="black",linetype=2)+
	labs(x="Concentration A",
		 y="Concentration B",
		 fill="Effect",
		 title="Protective Surface (linear)")+
	coord_equal()
```

This simple manipulation produces a surface that is qualitatively very 
different from the original.  It might be described as a surface in which one
drug (the first, in this case) produces an inhibitory effect, dropping the
measured value from 1 to 0; and in which the second drug, though having no 
effect of its own, inhibits or protects against the effects of the first.  We
call this a "protective" surface.

Another qualitative shift can be observed when the surface is flipped along
both dimensions.  The resulting surface shows no effect from either drug alone,
but shows a pronounced effect when both compounds are present, a behavior we
refer to as "coactive".

```{r}
linearSurface <- linearSurface %>%
	mutate(coactive=evalFlippedBraidModel(concA,concB,
											c(1,1,3,3,0,1,1,1,0),
											flip="both"))
logSurface <- logSurface %>%
	mutate(coactive=evalFlippedBraidModel(concA,concB,
											c(1,1,3,3,0,1,1,1,0),
											flip="both"))

ggplot(logSurface,aes(x=concA,y=concB)) +
	geom_raster(aes(fill=coactive)) +
	scale_fill_distiller(palette="RdYlBu",direction=1) +
	geom_contour(aes(z=coactive),breaks=seq(0.1,0.9,by=0.1),
				 colour="black",linetype=2)+
	scale_x_log10()+
	scale_y_log10()+
	labs(x="Concentration A",
		 y="Concentration B",
		 fill="Effect",
		 title="Coactive Surface (logarithmic)")+
	coord_equal()
ggplot(linearSurface,aes(x=concA,y=concB)) +
	geom_raster(aes(fill=coactive)) +
	scale_fill_distiller(palette="RdYlBu",direction=1) +
	geom_contour(aes(z=coactive),breaks=seq(0.1,0.9,by=0.1),
				 colour="black",linetype=2)+
	labs(x="Concentration A",
		 y="Concentration B",
		 fill="Effect",
		 title="Coactive Surface (linear)")+
	coord_equal()
```

Such atypical response surfaces, though less common than the classical model,
arise in a wide range of circumstances, and are just as critical to model and
quantify.  `braidrm` therefore includes several functions for evaluating,
manipulating, and fitting such surfaces.

## Protective Surfaces

The included data object `protectiveExample` includes surface in which drug A
produces a prounounced and potent effect, which is then suppressed at high
concentrations by the otherwise inactive drug B; a behavior we refer to as
"protective".  The function `fitProtectiveBraid_A()` fits a model in which drug
A is active, but inhibited by compound B.  We see that the resulting parameters
summarize the behavior seen, and the resulting fit surface matches the data
closely.

```{r, warning=FALSE}
measured <- protectiveExample

ggplot(measured,aes(x=concA,y=concB)) +
	geom_braid(aes(fill=measure))+
	scale_fill_distiller(palette="RdYlBu",direction=-1) +
	scale_x_log10()+
	scale_y_log10()+
	labs(x="Concentration A",
		 y="Concentration B",
		 fill="Effect",
		 title="Measured Surface")+
	coord_equal()

protectiveFit <- fitProtectiveBraid_A(measure ~ concA + concB, measured,
									  getCIs=FALSE)

measured$fit <- fitted(protectiveFit)

ggplot(measured,aes(x=concA,y=concB)) +
	geom_braid(aes(fill=fit))+
	scale_fill_distiller(palette="RdYlBu",direction=-1) +
	scale_x_log10()+
	scale_y_log10()+
	labs(x="Concentration A",
		 y="Concentration B",
		 fill="Effect",
		 title="Fit Surface")+
	coord_equal()

coef(protectiveFit)
```

## Oppositional Surfaces

The `oppositionalExample` object contains a response surface in which drug A
produces an excitatory effect (from 0 to 1), while drug B produces an effect in
the opposite direction. Though the effect of drug A dominates at higher
concentrations, the surface does not conform to the traditional assumptions of
the BRAID model.  The function `fitOppositionalBraid_A()` fits a flipped model
in which the two drugs produce effects in opposite directions, with the effect
of drug A taking over at the highest concentrations.

```{r, warning=FALSE}
measured <- oppositionalExample

ggplot(measured,aes(x=concA,y=concB)) +
	geom_braid(aes(fill=measure))+
	scale_fill_distiller(palette="RdYlBu",direction=-1) +
	scale_x_log10()+
	scale_y_log10()+
	labs(x="Concentration A",
		 y="Concentration B",
		 fill="Effect",
		 title="Measured Surface")+
	coord_equal()

oppositionalFit <- fitOppositionalBraid_A(measure ~ concA + concB, measured,
									  getCIs=FALSE)

measured$fit <- fitted(oppositionalFit)

ggplot(measured,aes(x=concA,y=concB)) +
	geom_braid(aes(fill=fit))+
	scale_fill_distiller(palette="RdYlBu",direction=-1) +
	scale_x_log10()+
	scale_y_log10()+
	labs(x="Concentration A",
		 y="Concentration B",
		 fill="Effect",
		 title="Fit Surface")+
	coord_equal()

coef(oppositionalFit)
```

## Coactive Surfaces

Finally, the `coactiveExample` object depicts a response surface in which
neither drug produces a noticeable effect in isolation, but the combined effect
when both are present is very pronounced.  Such "coactive" surfaces can be fit
by two functions: `fitCoactiveBraid_pure()` which holds that neither agent has
any measurable effect in isolation, and `fitCoactiveBraid_partial()`, which
allows either agent to have small partial effects.

```{r, warning=FALSE}
measured <- coactiveExample

ggplot(measured,aes(x=concA,y=concB)) +
	geom_braid(aes(fill=measure))+
	scale_fill_distiller(palette="RdYlBu",direction=-1) +
	scale_x_log10()+
	scale_y_log10()+
	labs(x="Concentration A",
		 y="Concentration B",
		 fill="Effect",
		 title="Measured Surface")+
	coord_equal()

coactiveFit <- fitCoactiveBraid_pure(measure ~ concA + concB, measured,
									  getCIs=FALSE)

measured$fit <- fitted(coactiveFit)

ggplot(measured,aes(x=concA,y=concB)) +
	geom_braid(aes(fill=fit))+
	scale_fill_distiller(palette="RdYlBu",direction=-1) +
	scale_x_log10()+
	scale_y_log10()+
	labs(x="Concentration A",
		 y="Concentration B",
		 fill="Effect",
		 title="Fit Surface")+
	coord_equal()

coef(coactiveFit)
```
